---
title: "01 – Loading Data and Preprocessing"
author: "Blake Hanson, PhD"
date: "`r Sys.Date()`"
output: html_document
---


```{r 01_setup, include=FALSE}
# Global chunk options for the workshop
knitr::opts_chunk$set(
  echo = TRUE,       # show code
  message = TRUE,    # show messages (helpful for learning)
  warning = FALSE    # keep output clean; we can turn warnings on if needed
)  

# Knit from the project root so relative paths like "data/..." work
knitr::opts_knit$set(root.dir = "..")
```

```{r packages_for_knit, include=FALSE}
# Source package installation/loading script (used for knitting only)
source("scripts/00_install_packages.R")
```

# Goal for this section

In this notebook, we will load the workshop data files and combine them into a single **phyloseq object** for analysis.

By the end of this notebook you will be able to:

1. Load the workshop data tables into R
2. Combine them into a single **phyloseq object**
3. Verify the phyloseq object is correctly formatted

> **Tip for Posit Cloud:** Run each code chunk in order using the green “Run” button.


## Step 1 — Confirm you are in the project root

Because we are working from a GitHub repository (and running in Posit Cloud), we rely on relative file paths rather than absolute paths (e.g., ~/Documents/...).

Relative paths only work if your working directory is set to the project root.

If this is not the case, R will not be able to find the input data files.

This chunk checks that the data/ folder exists in your current working directory and provides a helpful error message if it does not.

```{r check_data_repository}
# If/else statement that checks for the data/ folder and provides an informative error if not found. 
# Indicates the folder is found if present.
if (!dir.exists("data")) {
stop(
"The 'data/' directory was not found.\n",
"Make sure you opened the project (not just the script)."
)
} else {
message("Data directory found")
}
```


## Step 2 — Define paths to the workshop files

Rather than typing file paths repeatedly, we store them as named variables.

This improves readability, reduces errors, and makes scripts easier to modify or reuse.

> **Tip:** This is a good coding habit because:
>
> - It makes code more readable
> - It allows easy switching between relative and absolute paths if needed
> - It supports reproducible research and code sharing

```{r data_paths}
# Set a variable including the name of the folder that includes the data (checked for above)
data_dir <- "data"

# Path for the OTU file
otu_path      <- file.path(data_dir, "otu_table.csv")
# Path for the taxonomy file
taxonomy_path <- file.path(data_dir, "taxonomy_table.csv")
# Path for the metadata file
metadata_path <- file.path(data_dir, "metadata_table.csv")
```

To confirm that these variables contain the paths we expect, we print them to the console.

> **Tip:** Printing paths is another good habit — it helps with debugging, file tracking, and transparency.

```{r print_paths}
# Print the OTU file path
message("The OTU file path is: ", otu_path)
# Print the taxonomy file path
message("The taxonomy file path is: ", taxonomy_path)
# Print the metadata file path
message("The metadata file path is: ", metadata_path)
```


## Step 3 — Load the files into the R environment 

We now load the three core input tables used in microbiome analyses:

- OTU table: raw read/amplicon counts (taxa x samples)
- Taxonomy table: taxonomic labels for each OTU
- Metadata table: sample-level information used for grouping and comparisons

These tables are read from CSV files and stored in R as data frames. 

```{r load_data}
### Loading the dataset into R
# OTU table - this data is samples as columns and OTUs as rows
otu_df <- read.csv(otu_path, header = TRUE, row.names = 1, check.names = FALSE, stringsAsFactors = FALSE)
# Taxonomy table - the taxonomy for each OTU
taxonomy_df <- read.csv(taxonomy_path, header = TRUE, row.names = 1, check.names = FALSE, stringsAsFactors = FALSE)
# Metadata table - the metadata by sample
metadata_df <- read.csv(metadata_path, header = TRUE, row.names = 1, check.names = FALSE, stringsAsFactors = FALSE)
```


## Step 4 — Data formatting 

The phyloseq package expects data to be stored in specific object types rather than plain data frames.

In this step, we convert each table into the appropriate phyloseq format:

- otu_table() for count data
- tax_table() for taxonomy
- sample_data() for metadata

```{r data_formatting}
## Format each data frame for phyloseq
# Format OTU data frame
phyloseq_otu  <- otu_table(as.matrix(otu_df), taxa_are_rows = TRUE)
# Convert taxonomy table into a matrix and format
taxonomy_matrix <- as.matrix(taxonomy_df)
phyloseq_taxonomy  <- tax_table(taxonomy_matrix)
# Format metadata data frame
phyloseq_metadata <- sample_data(metadata_df)
```


## Step 5 - Combine the formatted objects into a phyloseq object 

phyloseq is designed to keep counts, taxonomy, and metadata synchronized.

The core object types we just created are:

- otu_table()
- tax_table()
- sample_data()

We now combine them into a single phyloseq object, which will be used throughout the rest of the workshop.

```{r create_phyloseq_object}
# Combine the formatted objects into a phyloseq object
phyloseq_object <- phyloseq(phyloseq_otu, phyloseq_taxonomy, phyloseq_metadata)
```


## Step 6 — Quick checks (what is inside the phyloseq object?)

Before moving on, it’s important to verify that everything loaded correctly.

We start by printing a summary of the phyloseq object.

```{r check_phyloseq_object}
phyloseq_object
```
Next, we inspect the first few rows of each component.

```{r head_sample}
head(sample_data(phyloseq_object))
```

```{r head_taxonomy}
head(tax_table(phyloseq_object))
```

```{r head_otu}
head(otu_table(phyloseq_object))
```

# Summary

At this point you have a working `phyloseq_object` that contains:

* OTU counts (`otu_table`)
* taxonomy labels (`tax_table`)
* sample metadata (`sample_data`)

This object ensures that all data types remain aligned as we subset, transform, analyze, and visualize the microbiome data. 

In the next notebook, we will use this object to:

- Describe the microbiome using taxonomic summaries
- Calculate alpha diversity
- Generate figures that render in Posit Cloud
